\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}

### 2014-06-19 paper/semantics.tex
As per discussion with Carlo today, beginning to tighten up the text at the beginning of section 3.  The goal is to be more concise, as the reader need not be handheld through staging itself

Every expression in \lang can be understood as having a {\em stage},
either \bbone~or \bbtwo, in addition to a type.
Just as a type describes \emph{which} values an expression reduces to,
the stage tells us \emph{when} it will be fully reduced. 
% FIXME why?
The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.

The only types that exist at stage \bbtwo{} are products, sums, functions, and base types.
In addition to this usual set, stage \bbone{} also has one more type: $\fut A$,
where $A$ is a stage-\bbtwo{} type.
This can be understood as \emph{promising} a value of type $A$ at stage \bbtwo.
We'll see shortly that the promise given by $\fut A$ can only be treated opaquely at stage one; 
it cannot be redeemed for an actual $A$ until the second stage.

Although there is a strict one-way dependence between the two stages at the type
level, their terms are defined mutually recursively:
$\next$ embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with \{braces\} in the concrete syntax to clearly indicate the stage boundaries with in a program.
The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.

%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.

The grammar and type system for our language, \lang, are given in
\ref{fig:grammar,fig:statics}.
We annotate the typing judgment and context variables with stages; only $\fut$
and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.

Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
This promise can only be redeemed be the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.

The $\pause$ construct is a mechanism to wrap stage \bbone~integers so that they can be used at stage \bbtwo.  
Although $\pause$ can be implemented from our other features, 
we provide it as a primitive to simplify our examples.

@@

Fast exponentiation is a method for calculating $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

Ignoring for the moment how to define recursive functions in \lang, we can
implement staged fast exponentiation as follows:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}

>>>>

In this section, we describe \lang, a simple two-stage language
used as the basis for defining and analyzing stage splitting.
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
Every expression in \lang\ has a statically known {\em stage},
either \bbone~or \bbtwo, which defines when it is fully reduced.
%\footnote{Typing judgments and context variables are annotated with stages.}
%
% in addition to a type.
% Just as a type describes \emph{which} values an expression reduces to,
%the expression's stage tells us \emph{when} it will be fully reduced. 
% FIXME why?
%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
Stage-\bbone{} and stage-\bbtwo{} types include products, sums, functions, and base types.
In addition to these types, stage-\bbone{} expressions may also be of type $\fut A$,
representing an enclosed stage-\bbtwo{} expression of type $A$.
$\fut A$ terms must be treated opaquely by stage-\bbone{} code
as they cannot be reduced (to values with type $A$) until stage \bbtwo.

Although there is a strict one-way dependence between the two stages at the type
level, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (The figures annotate typing judgments and context variables with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
Although it is possible to implement $\pause$ from other \lang\ features, 
we provide it as a core primitive to simplify our examples.

% As our concern is stage splitting of explicitly staged programs expressed in \lang.
We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent spitting.

%The problem of how to convert an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
%We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.

@@

As a simple example of expressing staged programs in \lang\, consider the following fast exponentiation algorithm below, which 
calculates $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The algorithm is expressed in \lang\ below:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}
